---
title: 'SQL Intro: Querying Databases'
author: "Kaymal"
output:
  github_document: default
---

Structured Query Language (SQL) is a language for interacting with data stored in a _relational database_. The focus of this notebook is **querying databases**, yet one can also use SQL to modify a database or to create one. [Sakila](https://www.google.com/url?sa=t&rct=j&q=&esrc=s&source=web&cd=1&cad=rja&uact=8&ved=2ahUKEwipzNLVifHgAhURqaQKHXPWAfoQFjAAegQIBhAB&url=https%3A%2F%2Fdev.mysql.com%2Fdoc%2Fsakila%2Fen%2F&usg=AOvVaw2NuLo1XbYVnyU8rBTLsipX) Database is used in this study.

Outline:

* Installing Necessary Packages
* Connection and Credentials
* Basic SQL Commands
    + Viewing the Data
    + Selecting Data
        - SELECT
        - DISTINCT
        - LIMIT
        - COUNT
    + Filtering Results
        - WHERE
        - WHERE AND/OR
        - BETWEEN
        - WHERE IN
        - IS NULL/IS NOT NULL
        - LIKE/NOT LIKE
    + Aggregate Functions
        - COUNT
        - AVG, MAX, MIN, SUM
        - WHERE + Aggregate Function
    + Sorting, Grouping, Merging Data
        - ORDER BY
        - GROUP BY
        - HAVING

## Installing Necessary Packages 

```{r install packages}

#install.packages("RMariaDB")
library(DBI)
library(RMariaDB)
library(RMySQL)

library(knitr)
```
```{r config, eval=FALSE}
# Install config for credentials
# install.packages("config")
library(config)
```

## Connection and Credentials

For security reasons, it is always better not to write username or passwords in a Notebook. Creating a `config.yml` file and reading in the credentials from that is one option for security.

```{r credentials}
db <- config::get("DB")
```


```{r connect}
# Initiate a connection with RMySQL

con = dbConnect(RMySQL::MySQL(),
      user = db$user, password = db$password, dbname = "sakila", port = 3306)

# Alternatively, initiate a connection with RMariaDB
#con = dbConnect(RMariaDB::MariaDB(),
#      user = db$user, password =db$password, dbname = "sakila", port=3306)

# Remove credentials
rm(db)
```

## Basic SQL Commands

### Viewing the Data

```{sql, connection = con}
-- Print names of the tables in the 'sakila' Database
SHOW TABLES;
```

### Selecting Data

#### SELECT
We can select data **FROM** a table using the **SELECT** keyword.

```{sql, connection=con}
SELECT 'First query'
AS first_result;
```


```{sql, connection = con, max.print = 5}
-- Select all columns from the 'film' table
SELECT *
FROM film;
```

```{sql, connection = con, max.print = 5}
-- Select the 'title' column from the 'film' table
SELECT title
FROM film;
```

```{sql, connection = con, max.print = 5}
-- Select the 'title' and 'release_year' columns from the 'film' table
SELECT title, release_year
FROM film;
```

#### DISTINCT

When we want to return the unique values in a column, we use **DISTINCT** keyword.

```{sql, connection = con, max.print = 5}
-- Get different types of 'rating' values. 
SELECT DISTINCT rating
FROM film;
```

#### LIMIT

We can use `LIMIT` to return a certain number of lines. 

```{sql, connection = con, max.print = 5}
-- Select all columns from the 'film' table. Limit the output with 2 results.
SELECT *
FROM film
LIMIT 2;
```

#### COUNT

```{sql, connection = con}
-- Get the number of rows in the 'film' table
SELECT COUNT(*)
FROM film;
```

```{sql, connection = con}
-- Print the number of distinct 'rating's in the 'film' table
SELECT COUNT(DISTINCT rating)
FROM film;
```

### Filtering Results

#### WHERE

**WHERE** is used to filter the results. We use comparison operators (`=`, `<=`, `<`, `<>`, etc.) along with WHERE. Note that we use `=` instead of `==`, and `<>` instead of `!=`.

```{sql, connection = con, max.print = 5}
-- Select all details for films which are released in 2006
SELECT *
FROM film
WHERE release_year = 2006;
```

#### WHERE AND/OR

```{sql, connection = con, max.print = 5}
-- Select all details for films which are released in 2006 and have rating 'G'
SELECT *
FROM film
WHERE release_year = 2006
AND rating = 'G';
```

If we want to combine `AND` and `OR`, we need to use paranthesis.

```{sql, connection = con, max.print = 5}
-- Select all details for films which are released in 2006 and have rating 'G' or 'R'
SELECT *
FROM film
WHERE release_year = 2006
AND (rating = 'G' OR rating = 'R');
```

#### BETWEEN

```{sql, connection = con}
-- Select all details from the 'table'sales_by_film_category' table where total sales are between 4500 and 5000.
SELECT *
FROM sales_by_film_category
WHERE total_sales
BETWEEN 4500 AND 5000;
```

#### WHERE IN

When we need to filter based on multiple values, we can use `WHERE IN` instead of multiple `OR`s.

```{sql, connection = con}
-- Select all details from the 'table'sales_by_film_category' for the specified categories.
SELECT *
FROM sales_by_film_category
WHERE category IN ('Animation', 'Action', 'Drama')
```

#### IS NULL/IS NOT NULL

`NULL` is particularly usefull when we want to learn whether there is a missing/unknown value in data.

```{sql, connection = con}
-- Select all details from the 'film_list' where 'category' is missing.
SELECT *
FROM film_list
WHERE category IS NULL;
```
```{sql, connection = con}
-- Count the number of missing values of the  'original_language_id' coulmn from the 'film' table.
SELECT COUNT(*)
FROM film
WHERE original_language_id IS NULL;
```

#### LIKE/ NOT LIKE

We can use `LIKE` and `NOT LIKE` for filtering by a _pattern_. Wildcard `%` is used to match zero or more characters, while `_` is used to match a single character.

```{sql, connection = con}
-- Select title from film where title starts with MAT.
SELECT title
FROM film
WHERE title LIKE 'MAT%';
```
```{sql, connection = con}
-- Select title from film where title's second and third characters are A and T respectively.
SELECT title
FROM film
WHERE title LIKE '_AT%';
```
### Agregate Functions
#### COUNT
#### AVG, MAX, MIN, SUM
#### WHERE + Aggregate Function

### Sorting, Grouping, Merging Data
#### ORDER BY
#### GROUP BY
#### HAVING
-----
```{r disconnect, eval=FALSE}
# Disconnect
dbDisconnect(con)
```
