---
title: 'SQL Intro: Querying Databases'
author: "Kaymal"
output:
  github_document: default
  html_document: default
---

Structured Query Language (SQL) is a language for interacting with data stored in a _relational database_. The focus of this notebook is **querying databases**, yet one can also use SQL to modify a database or to create one. [Sakila](https://www.google.com/url?sa=t&rct=j&q=&esrc=s&source=web&cd=1&cad=rja&uact=8&ved=2ahUKEwipzNLVifHgAhURqaQKHXPWAfoQFjAAegQIBhAB&url=https%3A%2F%2Fdev.mysql.com%2Fdoc%2Fsakila%2Fen%2F&usg=AOvVaw2NuLo1XbYVnyU8rBTLsipX) Database is used in this study.

Outline:

* Installing Necessary Packages
* Connection and Credentials
* Basic SQL Commands
    + Selecting
        - SELECT
        - SELECT DISTINCT
        - COUNT
    + Filtering
        - WHERE
        - WHERE AND/OR
        - BETWEEN
        - WHERE IN
        - NULL and ISNULL
        - LIKE and NOT LIKE
    + Aggregate Functions
    + Sorting, Grouping, Merging
        - ORDER BY
        - GROUP BY
        - HAVING

## Installing Necessary Packages 

```{r install packages}

#install.packages("RMariaDB")
library(DBI)
library(RMariaDB)
library(RMySQL)

library(knitr)
```
```{r config, eval=FALSE}
# Install config for credentials
# install.packages("config")
library(config)
```

## Connection and Credentials

For security reasons, it is always better not to write username or passwords in a Notebook. Creating a `config.yml` file and reading in the credentials from that is one option for security.

```{r credentials}
db <- config::get("DB")
```


```{r connect}
# Initiate a connection with RMySQL

con = dbConnect(RMySQL::MySQL(),
      user = db$user, password = db$password, dbname = "sakila", port = 3306)

# Alternatively, initiate a connection with RMariaDB
#con = dbConnect(RMariaDB::MariaDB(),
#      user = db$user, password =db$password, dbname = "sakila", port=3306)

# Remove credentials
rm(db)
```

## Basic SQL Commands

### SELECT

We can select data **FROM** a table using the **SELECT** keyword.

```{sql, connection=con}
SELECT 'First query'
AS first_result;
```


```{sql, connection = con}
-- Print tables of the 'sakila' Database
SHOW TABLES;
```

```{sql, connection = con}
-- Print the number of rows in the 'film' table
SELECT COUNT(*)
FROM film;
```

```{sql, connection = con, max.print = 5}
-- Select all columns from the 'film' table
SELECT *
FROM film;
```

```{sql, connection = con, max.print = 5}
-- Select the 'title' column from the 'film' table
SELECT title
FROM film;
```

```{sql, connection = con, max.print = 5}
-- Select the 'title' and 'release_year' columns from the 'film' table
SELECT title, release_year
FROM film;
```

```{r disconnect, eval=FALSE}
# Disconnect
dbDisconnect(con)
```
